\chapter{Benchmark of methods}
The performance of NLP solvers and frameworks is often tested on benchmark problems, like the CUTEst \cite{gouldCUTEstConstrainedUnconstrained2014} dataset. I have created my own dataset for this purpose.
As written on Ipopt website: Ipopt has many (probably too many) options \cite{ipopt_options} that can be adjusted, and that is true for all solvers. Some of differences in this benchmark may be due to suboptimal setup of the solver. Mesh refinement method in these tests is the h method with error threshold $10^-1$ and transcription method is Legendre-Gauss-Radau if not stated otherwise. Tests were performed on Ryzen 9 9950X using 32 cores by setting environment variable \texttt{OMP\_NUM\_THREADS} to 32 and MadNLP ran on Nvidia RTX 5070. This benchmark tests the time to solve and failure rate for all implemented vehicle variants on benchmarking tracks. Videos of some of the benchmarks are available on the YouTube playlist~\cite{slapsim_benchmark_playlist}.
\label{sec:benchmark}

\section{Benchmark of collocation methods}
This benchmark tests collocation methods. The lap-time seems to vary significantly on the same track and the same vehicle with different transcription method. This may mean that mesh refinement error tolerance was set too high.
\input{../sync/thesisTables/summary.tex}
\captionof{table}{Benchmark summary statistics by transcription variant.}
\label{tab:benchmark-summary}
\input{../sync/thesisTables/results.tex}
\captionof{table}{Benchmark results by track, vehicle model, and transcription variant.}
\label{tab:benchmark-results}

\section{Benchmark of linear solvers}
The performance increase going from mumps to ma97 is significant, however one test hit the iteration limit. In previously run benchmarks, which are not included here I used only 8 cores and the time to solve the problem was not decreased by ma97, that hints ma97 has better parallelization support.  The lap-time is same for all solvers. In this test I did not include PARDISO as my license was limited only to two cores. 
\input{../sync/thesisTables/linearSolvers/summary.tex}
\captionof{table}{Linear solver benchmark summary statistics.}
\label{tab:linsolver-summary}
\input{../sync/thesisTables/linearSolvers/results.tex}
\captionof{table}{Linear solver benchmark results by track, vehicle model, and transcription variant.}
\label{tab:linsolver-results}

\section{Benchmark of NLP solvers}
I have tested Ipopt with MUMPS and MadNLP with MUMPS and with CuDSS solver on GPU. MadNLP is significantly faster on average compared to Ipopt, but failed in one case. With madnlp-gpu variant the automatic differentiation was calculated on CPU using JuMP and the solver ran on GPU~\cite{madnlp_gpu_tutorial}, so it was not a 100\% GPU solution. The madnlp-gpu was slightly slower than CPU variant, it was significantly faster for some test cases and slower for others, this hints that maybe with proper solver settings it will be always faster.
\input{../sync/thesisTables/backends/summary.tex}
\captionof{table}{NLP solver / backend benchmark summary statistics (IPOPT-MUMPS, MadNLP-CPU, MadNLP-GPU).}
\label{tab:backend-summary}

\input{../sync/thesisTables/backends/results.tex}
\captionof{table}{NLP solver / backend benchmark results by track, vehicle model, and transcription variant.}
\label{tab:backend-results}


\section{Benchmark of derivatives computation}
This section compares two ways to provide the Hessian to Ipopt. Exact second-order derivatives obtained through JuMP's automatic differentiation are compared against Ipopt's built-in limited-memory quasi-Newton approximation. Exact Hessian was faster in all tested cases. The quasi-Newton approximation also failed to converge in time on some of the harder problems.
\input{../sync/thesisTables/hessian/summary.tex}
\captionof{table}{Hessian approximation benchmark summary statistics (exact vs.\ limited-memory quasi-Newton).}
\label{tab:hessian-summary}

\input{../sync/thesisTables/hessian/results.tex}
\captionof{table}{Hessian approximation benchmark results by track, vehicle model, and transcription variant.}\label{tab:hessian-results}



