\chapter{Frameworks for Optimal Control Problem Formulation}
Process of solving an optimal control problem has 3 parts. Problem definition,
solving nonlinear program and optional mesh refinement. This chapter describes frameworks in which the NLP problem can be defined and solved. Framework is: programming language + algebraic modeling tool + automatic differentiation tool + mesh refining algorithm. Solving optimal control problem can be split into 3 steps described in figure \ref{fig:ocp-pipeline}

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[
      node distance=1.4cm,
      block/.style={rectangle, draw, thick, font=\small, minimum width=4cm, minimum height=0.9cm, align=center, fill=white},
      exit/.style={ellipse, draw, thick, font=\small, minimum width=3cm, minimum height=0.8cm, align=center, fill=white}
    ]
    \node[block] (def) {1. Problem definition};
    \node[block, below=of def] (nlp) {2. NLP solve};
    \node[block, below=of nlp] (mesh) {3. Mesh refinement (optional)};
    \node[exit, below=of mesh] (sol) {Solution};
    \draw[->, thick] (def) -- (nlp);
    \draw[->, thick] (nlp) -- (mesh);
    \draw[->, thick] (mesh) -- node[right, font=\small]{converged} (sol);
    \draw[->, thick] let \p1=(mesh.east), \p2=(def.east) in
    (mesh.east) -- (\x1+1.5cm,\y1) -- node[right, font=\small, align=left]{needs\\ refinement} (\x1+1.5cm,\y2) -- (def.east);
  \end{tikzpicture}
  \caption{Optimal control problem solving pipeline. Adapted from \cite[p. 91]{betts}}
  \label{fig:ocp-pipeline}
\end{figure}

The consensus in optimal control community about tools for problem formulation
is that:
\begin{quote}
  There is no shortage of optimal control software based on pseudospectral approaches. There are a number of other optimal control libraries that tackle similar kinds of problems, such as OTIS4, GPOPS-II, and CasADi. \cite{Falck2021,dymos2024}
\end{quote}


\section{CasADi}
CasADi is an open-source tool for nonlinear optimization and algorithmic
differentiation.\cite{CasADi}\cite{Andersson2019}.
It interfaces with C++, Python, MATLAB. It provides a framework to build
the problem, and calculate the Hessian and Jacobian for the NLP solver. It does not
have any mesh refinement algorithm. Commonly paired with Ipopt, but is compatible with large variety of solvers.

\section{OpenMDAO + Dymos}
OpenMDAO \cite{Gray2019a} was developed at NASA
Glenn research center in python for evaluating aircraft concepts. Dymos implements hp mesh refinement and Lobatto and Radau collocation as default \cite{Falck2021}. According to forum post from Julia
Discourse, which may be biased towards Julia it is slower compared to InfiniteOpt
in Julia \cite{DoesAnybodyKnow2024}. It is the only open source tool I found that offers mesh refinement.
\section{Acados}
Acados is a modular open-source framework for fast embedded optimal control, developed at University of Freiburg~\cite{Verschueren2021}. It has C, Python, MATLAB + Simulink and Octave interface. This framework supports generating C code. It is demonstrated by numerous real-world examples, such as drone racing with \emph{replanning}, robot navigation in cluttered environment and many others \cite{acados_real_world_examples}.

\section{GPOPS-II}
GPOPS-II \cite{pattersonGPOPSIIMATLABSoftware2014} is perhaps the most advanced
tool for optimal control, it was developed at University of Florida by Michael
Patterson and Anil V. Rao. Other tools allow defining the problem, but
GPOPS-II is a level higher. It also includes advanced mesh refinement
strategies, multi-phase OCP and automatic scaling. For solving the problem
it uses Ipopt or SNOPT. It interfaces with MATLAB and C++, uses ADigator
\cite{weinstein2017algorithm} for automatic differentiation. This tool is
commercial and closed source. Implementation in this thesis is heavily based on
work of community from University of Florida.
\section{Julia + JuMP + InfiniteOpt / ExaModels}
JuMP is the go-to tool for algebraic modeling in Julia, it is mature and
versatile. Supports Linear programming, quadratic programming, mixed linear
programming, Nonlinear programming. ExaModels \cite{shin2023accelerating} is a
specialized algebraic modeling tool, it utilizes Single instruction, multiple
data methods, enabling derivative evaluation on GPU accelerators. Coupled with
MadNLP solver, the entire computation of nonlinear optimization can be run on a
GPU. No Julia package has mesh refinement. Julia offers several framework options
\begin{enumerate}
  \item Julia + JuMP
  \item Julia + JuMP + InfiniteOpt
  \item Julia + ExaModels
\end{enumerate}
I chose Julia + JuMP,


%\begin{table}[h]
%  \centering
%  \small
%  \begin{tabular}{|p{4.5cm}|p{4.5cm}|p{4.5cm}|}
%    \hline
%    \textbf{Aspect} & \textbf{Julia + JuMP + InfiniteOpt} & \textbf{Python + OpenMDAO + dymos} \\
%    \hline
%    Language / runtime
%      & Julia, JIT-compiled, no GIL
%      & Python, interpreted, GIL (NumPy releases) \\
%    \hline
%    License
%      & MIT (all components)
%      & Apache 2.0 \\
%    \hline
%    Algebraic modeling style
%      & Macro-based (\texttt{@variable}, \texttt{@constraint}, \texttt{@objective})
%      & Component graph (\texttt{ExplicitComponent}, \texttt{Group}, promotes/connects) \\
%    \hline
%    Derivatives
%      & AD via MathOptInterface; sparsity from expression graph
%      & MAUD unified derivatives; mix analytic / AD / FD / complex-step per component, framework chains them \\
%    \hline
%    Transcription methods
%      & Finite difference (Forward, Central, Backward); orthogonal collocation on finite elements with Gauss--Lobatto nodes only
%      & Radau pseudospectral, Gauss--Lobatto, explicit shooting, ExplicitShooting RK \\
%    \hline
%    Mesh / grid refinement
%      & None built-in; user must re-solve on refined supports
%      & Built-in ph- and hp-adaptive grid refinement \\
%    \hline
%    Multi-phase trajectories
%      & Manual; user concatenate \texttt{InfiniteModel}s and link
%      & Native \texttt{Trajectory} and \texttt{Phase} with linkage constraints \\
%    \hline
%    Multidisciplinary coupling
%      & Not supported; pure infinite-dim model
%      & Native MDO coupling with implicit/explicit components, cyclic groups, nonlinear/linear sub-solvers \\
%    \hline
%    NLP solvers
%      & IPOPT, MadNLP, Knitro, SNOPT and any MOI-compatible solver
%      & IPOPT, SNOPT, scipy via pyOptSparse \\
%    \hline
%    GPU acceleration
%      & Yes via ExaModels + MadNLP (SIMD on GPU)
%      & Not native \\
%    \hline
%    Parallelism
%      & Native multithreading, Distributed, MPI
%      & MPI across components \\
%    \hline
%    Sparse total Jacobian
%      & Yes, from MOI expression graph
%      & Yes, MAUD assembles sparse total Jacobian from component partials \\
%    \hline
%    Maturity / community
%      & JuMP mature in OR / energy; InfiniteOpt newer, smaller user base
%      & NASA-backed, large aerospace user base \\
%    \hline
%    Learning curve for OCP only
%      & Low; close to mathematical formulation
%      & Higher; OpenMDAO component model required even for single-discipline OCP \\
%    \hline
%  \end{tabular}
%  \caption{Comparison of Julia (JuMP + InfiniteOpt) and Python (OpenMDAO + dymos) stacks for optimal control problem formulation.}
%  \label{tab:julia-vs-python-ocp}
%\end{table}

