\chapter{NLP solvers,GPU exploitation???}
Once the objective function and constraints are created, the problem is passed
to NLP solver. Very popular methods are sequential quadratic solvers and
interior point solvers.
\section{Interior point methods: IPOPT, MadNLP}
\label{sec:interior_point}

Interior point solvers or barrier solvers work by simplifying the NLP into a
barrier problem and solving a sequence of barrier problems, with decreasing
barrier parameter. Solver used in this thesis is IPOPT
\cite{wachter_implementation_2006}. Article in \cite{vanaret_implementing_2025}
describes framework that unifies algorithmic components of different solvers,
allowing for custom new solvers.

\subsection{Barrier problem formulation}
\label{subsec:barrier_problem_formulation}
There are multiple ways to transform NLP in to barrier problem, this section
will briefly describe how IPOPT does it.
IPOPT article~\cite{wachter_implementation_2006} Describes how to transform generic
NLP into form~\ref{eq:barrier_problem} $h(\mathbf{x}) \geq 0$, interior point
methods solve a barrier problem.

\begin{equation}
    \begin{aligned}
        \min_{\mathbf{x} \in \mathbb{R}^n} \quad & \phi_\mu(\mathbf{x}) := f(\mathbf{x}) - \mu \sum_{i=1}^{n} \ln(x_i^{(i)}) \\
        \text{s.t.} \quad                        & c(\mathbf{x}) = 0
    \end{aligned}
    \label{eq:barrier_problem}
\end{equation}

Then barrier problem is further transformed into set of linear equations, that
compute newton step.
\begin{equation}
    \begin{bmatrix}
        W_k   & A_k & -I  \\
        A_k^T & 0   & 0   \\
        Z_k   & 0   & X_k
    \end{bmatrix}
    \begin{bmatrix}
        d_x^k       \\
        d_\lambda^k \\
        d_z^k
    \end{bmatrix}
    = -
    \begin{bmatrix}
        \nabla f(x_k) + A_k \lambda_k - z_k \\
        c(x_k)                              \\
        X_k Z_k e - \mu_j e
    \end{bmatrix}
\end{equation}

There are numerous commercial and free algorithms to solve this. For smaller
systems it is LU decomposition, but for larger sparse problems a dedicated
solver is a must. IPOPT is most commonly used with MUMPS \cite{MUMPS:1} solver
which is free. There are several linear commercial solvers, Harwell Subroutine
Library(HSL) supplies ma27 ma87 ma97 \cite{hsl2011collection} pardiso solver
\cite{pardiso}, Harwell Subroutine Library solvers are available free for
academic usage and pardiso solver is available free for academic but only
single core.

\section{Sequential quadratic programming}

\section{Hessian calculation trough Algorithmic differentiation}
Mayor complication of many NLP solvers is that hey require Hessian of the
objective function in order to compute the step. computation of it can be time
intensive. Most universal method is finite difference, that means evaluating
the function around current point and estimating the hessian. other method is
using symbolic differentiation, often the full symbolic expression is too large
to be evaluated effectively due to chain and product rule. Algorithmic
differentiation aims to solve problems of previous methods, by appling the
chain rule to the elementary operations of a function. Comprehensive review of
automatic differentiation is in article \cite{cortildBriefReviewAutomatic2023}

\section{Scaling and matrix conditioning}
Common pitfall of nonlinear programming is poor scaling, this happens when the
solver encounters variables, that are orders of magnitude apart, for example
motor force in thousands of newton and heading angle in radians. This happens,
because at the core of nonlinear programming solver is a linear solver
(MUMPS,MA97,PARDISO,...), that has to do matrix factorization every iteration,
to solve KKT linear system. If this linear system is ill conditioned due to
improper scaling, the factorization may be imprecise and search direction
corrupted.

Another thing which led to issues is constraints qualification. Linear
indpendence constiant qualification requires gradients of the active
constraints to be linearly independent, then the jacobian of the constraints is
rank deficient and linear solve may fails

\section{sparsity}
Sparsity is good, sparsity means matrix hessians and jacobians with many zeros
also sparsisty in constraints, that means not having super very long expression
with many terms in constraint. I guess it is beacause deriavtive is difficult
then and can lead solver in wrong direction or take small steps. Therefore
using more decison variables to define same problem can actually lead to faster
computation \cite{julia_discourse_ipopt_hessian}. Certain solvers can exploit the sprasity strcuture, such as FATROP\cite{vanroyeFATROPFastConstrained2023}. In source \cite{jorisgillisIPOPTSteroidsFatrop2024}  Joris gillis a  casadi developer compares IPOPT to FATROP, with  \textbf{FATROP being 33 times faster on a trajecotry optimisation problem.} 

Also \cite{betts} describes how sprasity structure helps with afster hessian aprprixmiation for using finite difference??? secant method.

\section{Initialization}
NLP solvers require initial guess for all decision variables, to start solving.
For this problem initial guess is obtained driving the vehicle on given track.
PD controller is used to keep the car on the centerline and P controller is
used to keep velocity at 5m/s. In general, it is possible to initialize all
variables at zero, but with large scale nonlinear programming this approach is
not used. For this problem initialization with zeros leads to singularity in
equation \ref{time2path} and is not possible at all. must be ineitialzied
otherwise very baaaadt?

\section{Global and local optimums}
This optimization problem is generally non convex, therefore any local optimum
does not guarantee global optimum.Commonly used NLP solvers, like IPOPT
guarantee only local optimums and not global ones. Good initization is basis
for finding a local optimum close to the global optimum. Source?????