\chapter{The Optimal Control Problem}
In previous chapter I have described how to build differential equations that goven movement of vehicle.
Methods discussed in this section are doing eesentialy the same thing, solving controlled ODE along a time/path trajectory. But their apporach and results are different.
The most common method for solving controlled ODE is time marching. Where the state computation is always for the states at time t+1, after solvig at t+1, time is moved/marched forward and solution a t +2 is calculated and so on. This gives a solution to ODE. If the ODE can be controlled like the car, you also need a separate controller, which becomes part of the ODE. This approach gives a solution to the problem of how to control car, so that it completes a lap around a circuit. However, this solution is not optimal to time or any objective as the controls depend on the controller. Therefore, if the car parameters change and the controller doesnt, the resulting time can be exatly same, becasue controller cannot take  advantage of lighter car. Sensitivity analysis and lap time can ba obtained but it is skewed by the controller chosen. In Car maker the ocntroller is driver and they have Driver adaptation. Using time marhcing is not hbest for sensitivity analysis a nd lap time simulation.
Controller can be tuned after each change of model but, that is time intensive. Another approach to solving controlled ODE is using shooting methods. Now the algorithm doesn't compute the states at only next time point, but it computes all states at all time points at the same time. Using this method it is possible to get rid of the controller, and instead of control rule a feed forward control vector along the trajectory is computed. This chapter deals with transcribing the continuos problem into discrete steps
\section{Nonlinear programming}

General NLP problems can be written as:
\begin{equation}
\begin{aligned}
\min_{\mathbf{x} \in \mathbb{R}^n} \quad & f(\mathbf{x}) \\
\text{s.t.} \quad & \mathbf{h}(\mathbf{x}) = 0 \\
& \mathbf{g}(\mathbf{x}) \leq 0
\end{aligned}
\label{eq:general_nlp}
\end{equation}

\subsection{KKT conditions}
alpha and omega of nolinear programming, lord praise the  KKT conditions.

\begin{equation}
\begin{aligned}
    \nabla f(\mathbf{x}) + \sum_{i=1}^{m} \lambda_i \nabla h_i(\mathbf{x}) + \sum_{i=1}^{p} \mu_i \nabla g_i(\mathbf{x}) &= 0 \\
    \mathbf{h}(\mathbf{x}) &= \mathbf{0} \\
    \mathbf{g}(\mathbf{x}) &\leq \mathbf{0} \\
    \mu_i &\geq 0, \quad i = 1, \ldots, p \\
    \mu_i g_i(\mathbf{x}) &= 0, \quad i = 1, \ldots, p
\end{aligned}
\end{equation}

\begin{equation}
    \begin{pmatrix}
        \nabla^2 \mathcal{L} & A^T \\
        A                    & 0
    \end{pmatrix}
    \begin{pmatrix}
        -\Delta x \\
        \lambda
    \end{pmatrix}
    =
    \begin{pmatrix}
        \nabla f(x) \\
        c(x)
    \end{pmatrix}
\end{equation}




\section{Solving controlled ODE}


\subsection{Time marching}
\subsection{Forward backward pass}



\section{Transcription methods}

\subsection{RK methods}
Most straightforward method to transcribe continuos optimal control problem into non linear program, is to use explicit Runge Kutta solving methods as equality constraints on the dynamics of vehicle. Using euler method, the constraints for the dynamics of vehicle:
\begin{equation}
	\mathbf{x}_{k+1} = \mathbf{x}_k + h \cdot f(\mathbf{x}_k, \mathbf{u}_k)
\end{equation}
This works, but it may be the worst possible transcription method, due to high error and instability of euler method \cite{SolvingOrdinaryDifferential1987}. Problems of euler method encountered in time marching simulation also translate to nonlinear programming. Implict RK methods can be used too, In Initial value problems they are used less often, as they require iterative solving. Optimal control problem is Boundary value problem and not initial value problem, therefore:

\begin{quote}
    \ldots for a boundary value problem \ldots any scheme becomes effectively implicit. Thus, the distinction between explicit and implicit initial value schemes becomes less important in the BVP context.
\end{quote}
\cite[p.~69]{Ascher1995}

For boundary value problem it makes no sense to use explicit Runge kutta methods, as they have inferior stability compared to implicit runge kutta metthods.\cite[p.~4]{raoSurveyNumericalMethods2010}

For Boundary value problems commonly implicit runge kutta method used is from LobattoIIIA family, but any other method can be used. Book \cite{betts} uses only these.

\subsection{Collocation methods}

\begin{quote}
The concept of collocation is old and universal in numerical analysis \ldots For ordinary differential equations it consists in searching for a polynomial of de-gree s whose derivative coincides ("co-locates") at s given points with the vectorfield of the differential equation \ldots
\cite[p.~224]{hairer}
\end{quote}
Job of collocation method is to find a polynomial that approximates the solution of BVP along independent variable(most commonly time). Collocation methods are a subset of implicit runge kutta methods.
The main distinction between runge kutta merthods and collocation. Is that the result of collocation is always polynomial, that does not hold for general Runge Kutta method.
Collocation methods are the core of this thesis, I have implemented  Gauss radau, gauss, gauss lobatto. Collocation methods differ in points at which the derivative of polynomial coincides ("co-locates) with the vector field of differential equation. Commonly used methods in optimal control are radau, lobatto, legendre. They can be written in integral and differential form\cite{garg_overview_nodate}. Collocatiom methods in this thesis are based on Framework described in\cite{gargUnifiedFrameworkNumerical2010}

\section{Mesh refinment}
\subsection{collocation integral form}
integral form is straight forward,  same as other unge kutta Methods
\subsection{collocation differential form}
differeanital form requires to create differentiation matrix of the aproximating polnomial, which contains the difference of the approximating polynomial.
\subsection{p adaptive methods}
p adaptive methods work by increasing the order of polynomial, until the error is below user defined threshold. Figure shows tht error along track is not distributed equaly. Error is localised around areas with large jump in controls. These areas need finer sampling. P methodsa cannot focus these specific points directly, they focus entire trajectory.
\subsection{h adaptive methods}
h methods are used when solution of optimal control problem is approximated by many intervals with low degree polynomial. These methods can decide to split segment into two smaller segments, they use different techniques to do this. Method described in \cite{betts} calculates error using ????? DOPLNIT. Another method described in \cite{jisongzhaoshuangliAdaptiveMeshRefinement} uses techniques from essentially nonoscillatory (ENO) interpolation \cite{Harten1997} to decide where to split and remove nodes.
\subsection{hp adaptive methods}
 hp adaptive methods are state of the art in mesh refinment. They are combination of increasing/decreasing mesh segments and incresing/decreasing order of aproximating polynomial insede segments. One of the first hp-method and commonly cited is  decribed in article~\cite{darbyHpAdaptivePseudospectral2011} from University of Florida. They have produces numerous papers on hp- adaptive mesh refinemnt~\cite{agamawiCGPOPSSoftwareSolving2020} \cite{liuAdaptiveMeshRefinement2018}. Recent work on mesh refinemnt for functions with kinks~\cite{wilkaAdaptiveHpPolynomialBased2025}.


