%\usepackage{amsmath,amssymb,amsfonts}

% arara: makeindex: { style: class/ctuthesis }
% arara: bibtex

% The class takes all the key=value arguments that \ctusetup does,
% and a couple more: draft and oneside
\documentclass[twoside]{ctuthesis}

\ctusetup{
%	preprint = \today,% footer timestamp
%	mainlanguage = english,
%	titlelanguage = czech,
	mainlanguage = english,
	otherlanguages = {english},
	title-czech = {Metody numerického optimálního řízení pro minimalizaci času na kolo },
	title-english = {Numerical Optimal Control \\ for Minimum Lap Time },
	subtitle-czech = {Cesta do tajů kdovíčeho},
	%subtitle-english = {Journey to the who-knows-what wondeland},
	doctype = M,
	faculty = F3,
	%department-czech = {Katedra matematiky},
	%department-english = {Department of Control Engineering},
	author = {Bc. Richard Ciglanský},
	supervisor = {doc. Ing. Zdeněk Hurák, Ph.D.},
	%supervisor-address = {Ústav X, \\ Uliční 5, \\ Praha 99},
	%supervisor-specialist = {John Doe},
	%fieldofstudy-english = {Control engineering},
	%subfieldofstudy-english = {Mathematical Modelling},
	%fieldofstudy-czech = {Matematcké inženýrství},
	%subfieldofstudy-czech = {Matematické modelování},
	keywords-czech = {NLP, laptime simulation, ipopt, skopirovat z bakalarky},
	keywords-english = {word, key},
	day = 22,
	month = 5,
	year = 2026,
%	specification-file = {ctutest-zadani.pdf},
%	front-specification = true,
%	front-list-of-figures = false,
%	front-list-of-tables = false,
%	monochrome = true,
%	layout-short = true,
}

\ctuprocess

\addto\ctucaptionsczech{%
	\def\supervisorname{Vedoucí}%
	\def\subfieldofstudyname{Studijní program}%
}

\ctutemplateset{maketitle twocolumn default}{
	\begin{twocolumnfrontmatterpage}
		\ctutemplate{twocolumn.thanks}
		\ctutemplate{twocolumn.declaration}
		\ctutemplate{twocolumn.abstract.in.titlelanguage}
		\ctutemplate{twocolumn.abstract.in.secondlanguage}
		\ctutemplate{twocolumn.tableofcontents}
		\ctutemplate{twocolumn.listoffigures}
	\end{twocolumnfrontmatterpage}
}

% Theorem declarations, this is the reasonable default, anybody can do what they wish.
% If you prefer theorems in italics rather than slanted, use \theoremstyle{plainit}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{note}
\newtheorem*{remark*}{Remark}
\newtheorem{remark}[theorem]{Remark}

% Abstract in Czech
\begin{abstract-czech}
Diplomova praca skuma metody riesenia minimum manevering time problemu pre cestne vozdila a implementuje ich. Práca delí tento problém na menšie podproblémy: vytvorenie modelu vozidla, zápis problému optimálneho riadenia, riešenie nelineárneho programu, riešenie systému lineárnych rovníc a citlivostná analýza. Pre tieto podproblémy existuje mnoho rôznych možností riešenia, práca tieto možnosti vysvetľuje a porovnáva. V práci sú implementované mettódy riešenia podproblémov ktoré som považoval za najužitočnejšie. Práca obsahuje aj benchmark vybraných možnosti riešenia.

\end{abstract-czech}

% Abstract in English
\begin{abstract-english}
Thesis researches and implements ways of solving minimum manuevering time problem (lap time simulation) for road vehicles. Thesis divides this problem into smaller subproblems, for which there are already mature sofware methods available. Subproblems are: racer car modelling, transcription of optimal control problem into NLP, solving of NLP, solving sets of linear equations and sensitivity analysis. Thesis implements  these methods and benchmarks their performance.
x\par x\par x\par x\par x\par x\par x\par x\par x\par x\par x\par x\par x


\end{abstract-english}

% Acknowledgements / Podekovani
\begin{thanks}
Děkuji ČVUT, že mi je tak dobrou \emph{alma mater}.
\end{thanks}

% Declaration / Prohlaseni
\begin{declaration}
Prohlašuji, že jsem předloženou práci vypracoval samostatně, a že jsem uvedl veškerou použitou literaturu.

V Praze, \ctufield{day}.~\monthinlanguage{title}~\ctufield{year}
\end{declaration}

% Only for testing purposes
\listfiles
\usepackage{lmodern}
\usepackage[pagewise]{lineno}
\usepackage{lipsum,blindtext}
\usepackage{mathrsfs} % provides \mathscr used in the ridiculous examples
\usepackage{bm}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{emoji}
\usepackage{fontspec}
\usetikzlibrary{positioning,calc,shapes.geometric,arrows.meta,backgrounds,babel}

\begin{document}

\maketitle

\input{chapters/introduction.tex}

\part{Theory}
\input{chapters/vehicle_modelling.tex}
\input{chapters/optimal_control_problem.tex}
\input{chapters/nlp_solvers.tex}

\part{Implementation of Optimal Control problem}
\input{chapters/frameworks_ocp.tex}
\input{chapters/my_implementation.tex}

\part{results}
\input{chapters/comparison.tex}
\input{chapters/conclusion.tex}

\appendix

\printindex

\bibliographystyle{plain}
\bibliography{ctutest/ctutest}

\end{document}